---
title: "Machine Learning Project"
---


**Your Name**: Naman Lahoti
**Your G Number**: G01340550



```{r warning = FALSE, message = FALSE}
# Suppress dplyr summarise grouping warning messages
options(dplyr.summarise.inform = FALSE)

library(tidyverse)
library(tidymodels)
library(plotly)
library(highcharter)
library(discrim)
library(klaR)

credit_card_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/credit_card_df.rds'))

credit_card_df
```



# Data Analysis

In this section, you must think of at least 5 relevant questions that explore the relationship between `customer_status` and the other variables in the `credit_card_df` data set. The goal of your analysis should be discovering which variables drive the differences between customers who do and do not close their account.

You must answer each question and provide supporting data summaries with either a summary data frame (using `dplyr`/`tidyr`) or a plot (using `ggplot`) or both.

In total, you must have a minimum of 3 plots (created with `ggplot`) and 3 summary data frames (created with `dplyr`) for the exploratory data analysis section. Among the plots you produce, you must have at least 3 different types (ex. box plot, bar chart, histogram, scatter plot, etc...)

See the [Data Analysis Project](https://gmubusinessanalytics.netlify.app/data-analysis-project.html){target="_blank"} for an example of a question answered with a summary table and plot.

**Note**: To add an R code chunk to any section of your project, you can use the keyboard shortcut `Ctrl` + `Alt` + `i` or the `insert` button at the top of your R project template notebook file.



# Question 1


**Question**:
What is the customer account status based on the marital status and what is the average expenditure based on marital status?

**Answer**:
It can be found that there is no such huge difference between the expenditure of married and divorced customers, where the average expense of married customers is 3797.010 and divorced customers are 3678.593 respectively. We can say that single customers have spent more when compared to the other two categories.

From the plot it can be observed that most of the customers having active accounts are married, however single customers also do have a large part between the customer's marital status. 

In conclusion, from the analysis we can say that there are more chances that customers who are single are more likely to maintain their account active when compared to the other set of customers.

```{r}
#Yearly average expenditure based on marital status
credit_card_df %>% group_by(marital_status) %>% 
  summarise(yearly_avg = mean(total_spend_last_year))


#Customer status and their marital status
ggplot(credit_card_df, aes(y=marital_status))+geom_bar(aes(fill =customer_status),position = "dodge")+theme_minimal()+ theme_classic()  +xlab("Count") + ylab("Marital Status") + ggtitle(" Customer Status by Marital Status" )+  labs(fill = "Customer Status")+ scale_fill_brewer(palette="Blues")

```



# Question 2


**Question**:
Is there any relationship between the Education of a customer and the customer status?

**Answer**:
From the analysis it has been observed that there are more customers who have completed their Master's degree followed by Associates, Bachelor's and then doctorate customers.

The plot depicts that more number of customers who hold a Masters degree have an active account followed by customers with associates degree. But, interestingly, it is also seen that more number of customers (126 customers) with doctorate degree have closed account meaning they are inactive. In conclusion, reason should be found out that why there are more number of inactive accounts with customers holding a doctorate/PhD degree.

```{r}
#Customers and education
credit_card_df%>%
  dplyr::select(education)%>%
  count(education)

#plot of customer status and their education
plot_ly(credit_card_df,x = ~education, color = ~customer_status)%>%
  layout(title = "Education and Customer status",
         xaxis =list(title = 'Education'), 
         yaxis = list(title = 'Count'))

```


# Question 3


**Question**:
Is there any correlation among the Total credit Amount Average, marital status and employment status with respect to the account status?

**Answer**:
From the data, it has been observed that divorced customers having a full-time job have the highest average credit amount among all the other customers. And customers who are divorced and do some part-time job are the ones with the lowest average credit amount.

It has been observed from the plots that customers who are single and married do not use their credit line more when compared to the divorced customers.

```{r}
credit_card_df %>% 
  group_by(marital_status, employment_status) %>%
  mutate(credit_usage = credit_limit * utilization_ratio) %>%
  summarise(Total_credit_Amount_Avg = mean(credit_usage))%>%
  arrange(desc(Total_credit_Amount_Avg))

credit_card_df %>% 
  group_by(marital_status, employment_status) %>% 
  mutate(credit_usage = credit_limit * utilization_ratio) %>%
  ggplot(aes(x = credit_usage, fill = marital_status)) +
       geom_histogram(color = "yellow", bins = 20) + 
       facet_wrap( ~ marital_status, nrow = 1)+xlab("Credit Usage") + ylab("Count of customers")


```



# Question 4


**Question**:
Which type of credit card customers were mostly inactive during the last year?

**Answer**:
From the plot, it is observed that the average total months inactive for each customer in the last 12 months of both types of customers is fluctuating (because it’s a discrete data) with some outliers in the inactive/closed customers distribution. It is noticed that customers who have a closed account have been more inactive when compared to the customers with an active account.

```{r}
credit_card_df %>%
  ggplot( aes(months_inactive_last_year, fill=customer_status))+
  geom_density(alpha=0.6)+
  geom_vline(aes(xintercept=mean(months_inactive_last_year)),color="black",
             linetype="dashed", size=1)+theme_light() + scale_fill_manual(values =
                                                                            c("red", "green"))+
  ggtitle("Credit customers who were mostly inactive during the last year")+
  xlab("Months Inactive last year")+ylab("Frequency")+ theme(plot.title = element_text(size=15)) 


```



# Question 5


**Question**:
Who had made more than 100 transactions in the last 12 months according to marital status and what is the total percentage of transactions during the last year?

**Answer**:

From the data it can be interpreted that married customers (97 customers) having an active account have made more than 100 transactions in the last 12 months. While on the other hand, only 14 customers who are divorced and have an active account have made more than 100 transactions in the last 12 months.

One way to interpret the separating measures from the boxplot chart is:

50% of customers who left credit card services had number of transactions in the last 12 months less than or equal to 43, remembering that the maximum number of transactions in the last 12 months of customers who left the service is 72. While the median of the people who remained with the card services is 71 transactions.

75% of customers who left credit card services had a number of transactions in the last 12 months equal to or less than 51. The third quartile of people who remained with card services is 82 transactions.



```{r}
#Transactions more than 100 in the last year
credit_card_df %>% 
  dplyr::select(transactions_last_year,customer_status,marital_status)%>%
  filter(transactions_last_year>100)%>%
  count(customer_status,marital_status)%>%
  arrange(desc(n))

#Boxplot for total percentage of transactions during the last year
hcboxplot(outliers = TRUE, x = credit_card_df$transactions_last_year, var = credit_card_df$customer_status, name = "Length") %>%
  hc_title(text = "Total transactions in the last 12 months in terms of percentages", align = "center") %>%
  hc_yAxis(title = list(text = "Number of transactions in the last 12 months")) %>%
  hc_add_theme(hc_theme_smpl()) %>% 
  hc_legend(align = "center")
```
**Bonus Visualization**:
#Cancellations by Marital Status, education levels and card type (Treemap graphs)


```{r}
#cancellations by marital status
credit_card_df %>%
  filter(customer_status == "closed_account") %>%
  count(marital_status) %>%
  hchart("treemap", hcaes(x = marital_status, value = n, color = n)) %>%
  hc_colorAxis(stops = color_stops(colors = viridis::inferno(10))) %>%
  hc_title(text = "Number of Cancellations By Marital Status", align = "center")

#Cancellations by education
credit_card_df %>%
  filter(customer_status == "closed_account") %>%
  count(education) %>%
  hchart("treemap", hcaes(x = education, value = n, color = n)) %>%
  hc_colorAxis(stops = color_stops(colors = viridis::inferno(10))) %>%
  hc_title(text = "Number of Cancellations By Educational Level", align = "center")

#Cancellations by Card Type
credit_card_df %>%
  filter(customer_status == "closed_account") %>%
  count(card_type) %>%
  hchart("treemap", hcaes(x = card_type, value = n, color = n)) %>%
  hc_colorAxis(stops = color_stops(colors = viridis::inferno(10))) %>%
  hc_title(text = "Number of Cancellations by Card Type", align = "center")

```


# Machine Learning


In this section of the project, you will fit **three classification algorithms** to predict the outcome variable,`customer_status`.

You must follow the machine learning steps below. 

The data splitting and feature engineering steps should only be done once so that your models are using the same data and feature engineering steps for training.

- Split the `credit_card_df` data into a training and test set (remember to set your seed)
- Specify a feature engineering pipeline with the `recipes` package
    - You can include steps such as skewness transformation, correlation filters, dummy variable encoding or any other steps you find appropriate
- Specify a `parsnip` model object
    - You may choose from the following classification algorithms:
      - Logistic Regression
      - LDA
      - QDA
      - KNN
      - Decision Tree
      - Random Forest
- Package your recipe and model into a workflow
- Fit your workflow to the training data
    - If your model has hyperparameters:
      - Split the training data into 5 folds for 5-fold cross validation using `vfold_cv` (remember to set your seed)
      - Perform hyperparamter tuning with a random grid search using the `grid_random()` function
      - Refer to the following tutorial for an example - [Random Grid Search](https://gmubusinessanalytics.netlify.app/lesson-08-r-tutorial.html#Hyperparameter_Tuning14){target="_blank"}
      - Hyperparameter tuning can take a significant amount of computing time. Be careful not to set the `size` argument of `grid_random()` too large. I recommend `size` = 10 or smaller.
      - Select the best model with `select_best()` and finalize your workflow
- Evaluate model performance on the test set by plotting an ROC curve using `autoplot()` and calculating the area under the ROC curve on your test data




# Model 1
#Logistic Model
```{r}

#Creating training and testing datasets by using split
set.seed(245)

#Create a data split object
credit_card_split<-initial_split(credit_card_df,prop=0.7,strata = customer_status)

#Create the training data
credit_card_training<-credit_card_split%>%
  training()

#Create the testing data
credit_card_testing<-credit_card_split%>%
  testing()

# Check number of rows in each dataset
nrow(credit_card_training)
nrow(credit_card_testing)

```
```{r}
# Specify feature engineering recipe
credit_card_recipe<-recipe(customer_status~., 
                          data = credit_card_training)%>% 
  step_YeoJohnson(all_numeric(), -all_outcomes()) %>% 
  step_normalize(all_numeric(), -all_outcomes()) %>% 
  step_dummy(all_nominal(), -all_outcomes())

# Train and transform credit_card_recipe
credit_card_recipe %>% 
  prep(training = credit_card_training) %>% 
  bake(new_data = NULL)

```
```{r}
# Train logistic model
logistic_model<-logistic_reg()%>%
  set_engine("glm")%>%
  set_mode("classification")


#Create a workflow
credit_card_wkfl<-workflow()%>%
  #Include model object
  add_model(logistic_model)%>%
  #Include recipe object
  add_recipe(credit_card_recipe)

#Train Logistic model
logistic_fit<-credit_card_wkfl%>%
  fit(data=credit_card_training)
```

```{r}
#Obtain class predictions
class_preds<-predict(logistic_fit,new_data = credit_card_testing,type = "class")

#Obtain estimated probabilities
prob_preds<-predict(logistic_fit,new_data = credit_card_testing,type = "prob")

#Combine test set results
credit_card_results <- credit_card_testing %>% 
  dplyr::select(customer_status) %>% 
  bind_cols(class_preds, prob_preds)

credit_card_results

```


```{r}
#Create a confusion matrix
credit_card_results%>%
  conf_mat(truth=customer_status,estimate=.pred_class)

#Creating a custom metrics function
credit_card_metrics <- metric_set(accuracy, sens, spec, f_meas, roc_auc)

credit_card_metrics(credit_card_results, 
           truth = customer_status, 
           estimate = .pred_class,
           .pred_closed_account)
```
```{r}
#Train model with last_fit()
credit_card_last_fit<-credit_card_wkfl%>%
  last_fit(customer_status~.,split=credit_card_split)

# View test set metrics
credit_card_last_fit %>% 
  collect_metrics()

# Collect predictions
last_fit_results <- credit_card_last_fit %>% 
  collect_predictions()

last_fit_results

```
```{r}
#PLot ROC Curve
last_fit_results %>% 
  roc_curve(truth = customer_status, estimate = .pred_closed_account) %>% 
  autoplot()
```




# Model 2
#K-Nearest Neighbours(KNN)
```{r}
#set seed
set.seed(245)
#Dividing dataset into k-groups or folds each equal size
credit_card_folds <- vfold_cv(credit_card_training, v = 4)

#Create a KNN model
credit_card_knn_model <- nearest_neighbor(neighbors = tune()) %>% 
            #set engine to knn
             set_engine('kknn') %>% 
            #set mode as classification
             set_mode('classification')
```

```{r}
#create a workflow
credit_card_knn_wkfl<-workflow() %>% 
          #Include model object
          add_model(credit_card_knn_model) %>% 
          #Include recipe object
          add_recipe(credit_card_recipe)

## Create a grid of hyperparameter values to test
k_grid <- tibble(neighbors = c(10, 20, 30, 50, 75, 100, 125, 150))

#View Grid
k_grid

## Tune  workflow
set.seed(360)

knn_tuning <- credit_card_knn_wkfl %>% 
              tune_grid(resamples = credit_card_folds,
                         grid = k_grid)
```
```{r}
## Show the top 5 best models based on roc_auc metric
knn_tuning %>% show_best('roc_auc')

## Select best model based on roc_auc
best_k <- knn_tuning %>% 
          select_best(metric = 'roc_auc')

## View model
best_k

## Finalize workflow by adding the best performing model
final_knn_wkfl <- credit_card_knn_wkfl %>% 
                finalize_workflow(best_k)
```
```{r}
#Train and evaluate with last_fit()
last_fit_knn <- final_knn_wkfl %>% 
                last_fit(split = credit_card_split)

last_fit_knn %>% collect_metrics()
```


```{r}
#Collect Predictions
credit_card_knn_predictions <- last_fit_knn %>% 
                   collect_predictions()

credit_card_knn_predictions
```
```{r}
#Create a confusion matrix
conf_mat(credit_card_knn_predictions, truth = customer_status, estimate = .pred_class)
```


```{r}
#Plot ROC Curve
credit_card_knn_predictions %>% 
  roc_curve(truth = customer_status, estimate = .pred_closed_account) %>% 
  autoplot()


#F1 score
f_meas(credit_card_knn_predictions, truth = customer_status, estimate = .pred_class)
```




# Model 3
#Quadratic Discrminant Analysis(QDA)
```{r}
#QDA Model Specification
qda_model <- discrim_regularized(frac_common_cov = 0) %>% 
             set_engine('klaR') %>% 
             set_mode('classification')
```

```{r}
#Create a workflow
qda_wkfl <- workflow() %>% 
          #Add model object
          add_model(qda_model) %>% 
          #Add recipe object
          add_recipe(credit_card_recipe)
```

```{r}
#Train and evaluate with last_fit()
last_fit_qda <- qda_wkfl %>% 
                last_fit(split = credit_card_split)

#Collect Metrics
last_fit_qda %>% collect_metrics()

#Collect Predictions
qda_predictions <- last_fit_qda %>% 
                     collect_predictions()

qda_predictions

#F1 score
f_meas(qda_predictions, truth = customer_status, estimate = .pred_class)

```
```{r}
qda_predictions %>% 
  roc_curve(truth = customer_status, estimate = .pred_closed_account) %>% 
  autoplot()
```


# Summary of Results

Write a summary of your overall findings and recommendations to the executives at the bank. Think of this section as your closing remarks of a presentation, where you summarize your key findings, model performance, and make recommendations to improve customer retention and service at the bank.

Your executive summary must be written in a [professional tone](https://www.universalclass.com/articles/writing/business-writing/appropriate-tone-in-business-communications.htm){target="_blank"}, with minimal grammatical errors, and should include the following sections:

1. An introduction where you explain the business problem and goals of your data analysis

    - What problem(s) is this company trying to solve? Why are they important to their future success?
  
    - What was the goal of your analysis? What questions were you trying to answer and why do they matter?

<br>

2. Highlights and key findings from your Exploratory Data Analysis section 
    - What were the interesting findings from your analysis and **why are they important for the business**?

    - This section is meant to **establish the need for your recommendations** in the following section

<br>

3. Your “best” classification model and an analysis of its performance 
    - In this section you should talk about the expected error of your model on future data
      - To estimate future performance, you can use your model performance results on the **test data**
    - You should discuss at least one performance metric, such as an F1, sensitivity, specificity, or ROC AUC for your model. However, you must explain the results in an **intuitive, non-technical manner**. Your audience in this case are executives at a bank with limited knowledge of machine learning.

<br>

4. Your recommendations to the bank on how to reduce the number of customers closing their credit card accounts 
  
    - Each recommendation must be supported by your data analysis results 

    - You must clearly explain why you are making each recommendation and which results from your data analysis support this recommendation

    - You must also describe the potential business impact of your recommendation:
      
      - Why is this a good recommendation? 
      
      - What benefits will the business achieve?


**Summary**

Add your summary here. Please do not place your text within R code chunks.


*Introduction*
The credit_card_df data frame contains client information from a large U.S. bank which has the data of over 4000 customers having certain financial information and personal information like age, marital status, employment status, card type etc., The bank is looking to see if it can determine the factors that lead to customers canceling their credit card account and whether it can predict if a customer will cancel their account in the future.

*Problem Statement*
Large amount of customers leaving the credit card services of a bank is a major problem for the bank as it ruptures the credit line service and the profits of a reputed bank. Retention of customers is a major task that is to be carried out by banks and this is possible by observing the past data of the bank and the financial behavior of the customer based on the age and other crucial characteristics like dependents, income and the number of transactions made by the customer in the past.

To solve this problem, there is a need to analyze the root cause of why the customers are leaving the bank or being inactive for a larger period. Initially, we'd like to look at the company's financial losses using factor variables like education, marital status, employment status, and the card type they're using in relation to numeric variables like transactions made last year, total money spent last year, utilization ratio, and total accounts a customer has, as well as their impact on customer status.

This project is carried out in a sequence of steps, the first of which consists of an exploratory analysis, where the objective is to know the behavior of the variables and to analyze attributes that indicate a strong relationship with the cancellation of credit card service customers. Then the last step consists of applying a machine learning algorithm to find the best resources for building the model. 

*Key Findings*
The major discovery from the exploratory data analysis reveals several intriguing indicators that aid in determining the credit card status of clients. According to the research, there are a few aspects that have been identified as influential determinants in the customer's card status.

It can be found that there is no such huge difference between the expenditure of married and divorced customers, where the average expense of married customers is 3797.010 and divorced customers are 3678.593 respectively. We can say that single customers have spent more when compared to the other two categories.  Most of the customers having active accounts are married, however single customers also do have a large part between the customer's marital status. We can say that there are more chances that customers who are single are more likely to maintain their account active when compared to the other set of customers.

It has also been observed that there are more customers who have completed their master’s degree followed by Associates, Bachelor's and then doctorate customers. The insights define that a greater number of customers who hold a master’s degree have an active account followed by customers with associates degree. But interestingly, it is also seen that a greater number of customers (126 customers) with doctorate degree have closed account meaning they are inactive. In conclusion, reason should be found out that why there are a greater number of inactive accounts with customers holding a doctorate/PhD degree.

From the data, it has been observed that divorced customers having a full-time job have the highest average credit amount among all the other customers. And customers who are divorced and do some part-time jobs are the ones with the lowest average credit amount. It was analyzed from the plots that customers who are single and married do not use their credit line more when compared to the divorced customers.

The data also defines that the average total months inactive for each customer in the last 12 months of both types of customers is fluctuating (because it’s a discrete data) with some outliers in the inactive/closed customers distribution. It is noticed that customers who have a closed account have been more inactive when compared to the customers with an active account.

It is depicted that married customers (97 customers) having an active account have made more than 100 transactions in the last 12 months. While on the other hand, only 14 customers having marital status as divorced and have an active account have made more than 100 transactions in the last 12 months.

-------------------------------------------------------------------------------
Model    | Logistic Regression |       KNN      | QDA |
-------------------------------------------------------------------------------
Accuracy |     0.8711303    	 |   0.8538517    |   0.8833693  |
------------------------------------------------------------------------------
roc_auc  |     0.9437214       |   0.9307649 	|   0.9537380  |
-----------------------------------------------------------------------------
F1 score |     0.8430421      |   0.8364222 	|   0.8699839   |
-----------------------------------------------------------------------------


When comparing the performance of the above models, the Quadratic Discriminant Analysis (QDA) model appears to be the best, with the highest accuracy, the area under the ROC curve, and the F1 score.

The simplest intuitive performance metric is accuracy, which is just the ratio of properly predicted observations to all observations. This suggests that the analysis we conducted by forecasting the elements that influence consumers' decision to cancel their cards is 88% accurate with respect to other aspects, i.e., the factors that influence customer status. We can see how vital it is for the organization to keep track of, analyze, and improve these elements.


*Recommendations*

Self-employed consumers are problematic because they are more likely to cancel their credit card services. Since the number of cancellations were more in the blue card holders, schemes or discounts or low/no APR blue cards should be provided to retain such customers. It is also necessary to analyze such customers and understand the cause of blue card holders’ cancellation of accounts.

It is also advised to the bank to focus more on full-time working customers as they generate huge revenues for the organization by using the credit. Self-employed customers are problematic as they are more likely to cancel their credit card services.

By solving the above problems using the key findings and recommended solutions, the bank can function smoothly without any bottlenecks and there are also chances that customers might be retained.

Since retaining credit card customers is a necessity for banks, banks should initiate new schemes or offers to customers based on the customers portfolio and credit score. Accurate predictions are possible with the credit score of a particular customer.



